Comparison of two object-oriented languages: Eiffel and Ruby

### 2014-01-12 paper.tex
open classes

Ruby does not support multiple inheritance,  but a \emph{mixin} can be an equivalent feature. Before explaining what a mixin is it is important to explain modules in Ruby. A module is a sort of namespace grouping variables and functions together for obtaining a whole that provides functionalities. Modules cannot be instantiated, their purpose is to add functionality to a class. A mixin allows to include a module as a sort of superclass to the desired class, to mix the module in the class. It is possible to mix more than one module in a class, thus it looks very similar to multiple inheritance. In the example from figure \ref{fig:diamond}, the superclasses could be two modules that implement the different move behaviors. But even if Ruby uses mixins instead of multiple inheritance, the nameclash problem persists. Ruby resolves it automatically by overriding the previous definition, thus it is important for the programmer to be aware of this and give another name to one of the definition if the two method definitions are needed.

@@

Polymorphism is a key feature in object-oriented programming languages. Both Ruby and Eiffel support this feature but in a very different conceptual way. Eiffel supports subtype polymorphism, it means that Eiffel allows polymorphism only for the types that have a superclass in common. Thus inheritance is primordial for  subtype polymorphism. 
\\
\\ 
% http://en.wikipedia.org/wiki/Duck_typing
In Ruby it is also possible to achieve it using inheritance, but this is more a consequence of the mechanism that permits Ruby to support polymorphism. Ruby is dynamically typed and supports a special style of typing namely, \emph{duck typing}. Duck typing focusses on the methods of an object instead of its type. If the method is supported by the object it will be called whatever the type or output is. If the method call is not supported, then a run-time error is returned. Duck typing is the concept used for polymorphism in Ruby allows thus polymorphism without inheriting from a superclass. It is trivial why polymorphism also works with inheritance in Ruby, classes that inherit from a superclass inherit its methods and duck typing focusses on the presence of methods.
\\
\\
% http://blogs.msdn.com/b/ericwhite/archive/2008/10/29/anders-hejlsberg-introduces-c-4-0-at-pdc-2008.aspx
So even if both languages support polymorphism, their approach is completely different. Polymorphic calls are dependent of the type of objects in Eiffel and in Ruby they are dependent of the presence of the method. From a software engineering point of view it is logical that Eiffel focusses on the subtype polymorphism. First it is statically typed, thus there should be a specific type declared with the variable or parameter. However, this could be resolved with a keyword that instructs the compiler that a variable should be dynamically typed. Second and most importantly, this sort of solution is not in the philosophy of the Eiffel language. One of the goals of Eiffel is to produce software that is reliable and maintainable. If duck typing should be adapted in Eiffel then it would be only reliable if the programmers know exactly which type of objects will passed to the methods. The maintainability of the software would also be tricky, imagine that a method call changes of name in one of the classes, with duck typing it would not be clear that this method was used in a polymorphic call and by changing its name the programmer introduces errors that can be tedious to resolve. Thus it is important to be aware of the methods that are used in a polymorphic context when a language supports duck typing. Subtype polymorphism in a statically typed language is a much more reliable mechanism because the polymorphic calls are only possible for a restricted set of types and eliminate a lot of possible failures. It is also more easily maintained, because it is checked at compile time that every class implements the method. In Ruby's case, it is also logical that duck typing is used and not subtype polymorphism. First, subtype polymorphism is implicit with duck typing. Second, Ruby is dynamically typed, which means the types are known at run-time. This implies that there would be no secure way to ensure that the method call is applied on a particular class hierarchy. Perhaps it could be achieved with a kind of type cast for the desired superclass. But then it would still return a run-time error when the types are not correct, like it does when the method is not defined for an object. This is much more restrictive and goes also against the philosophy of Ruby which is simplicity.

@@

If there is one language specific feature that is essential in Eiffel, then this concept is \emph{Design by Contract}. The idea is that every system has interacting components and that their cooperation should follow some strict specifications (the contract) that settle the obligations and benefits for both client and supplier. The obligations have to be satisfied before feature calls. They are called \emph{preconditions} and are introduced by the \emph{require} keyword. The benefits describe what the result should be if the precondition was met. Benefits are thus \emph{postconditions} and can be specified using the \emph{ensure} keyword. Every contract also include \emph{class invariants}, which are conditions that have to be ensured during the lifetime of an object, including at its creation. Class invariants are specified after the \emph{invariant} keyword. These are the three main catergories of contracts and are implemented using assertions. Each assertion may be tagged, it is not mandatory and does not influence the contract but it is helpful for debugging and provides extra documentation. Since assertions are boolean expressions, it is possible to formulate them in function calls. This enables to express more complex conditions.

@@

There exist libraries in many programming languages that offers Design by Contract, even for \href{https://github.com/egonSchiele/contracts.ruby}{Ruby}.

@@

There are patterns that check if a variable is void-safe. The Certified Attachment Pattern (CAP) checks if a local variables or formal parameters is void. The attached syntax takes a step further. It is another sort of CAP that checks if the object is attached and provides a safe access to the objects that are attached as class attributes. Eiffel introduced two kind of types in order to assure the void-safeness of the softwarel:

@@

This is a feature that improves the reliability of the produced software and shows again that Eiffel's primary concern is to enhance the software quality.
\subsection{Ruby}
\subsubsection{Open Objects}
\subsubsection{Meta-class model}

@@

%DESIGN BY CONTRACT CODE
\lstinputlisting[label=e:contract,language=Eiffel,caption={Design by Contract in Eiffel}]{code_Eiffel/contract.e}

>>>>

Ruby does not support multiple inheritance,  but a \emph{mixin} can be an equivalent feature. Before explaining what a mixin is it is important to explain modules in Ruby. A module is a sort of namespace grouping variables and functions together for obtaining a whole that provides functionalities. Modules cannot be instantiated, their purpose is to add functionality to a class. A mixin allows to include a module as a sort of superclass to the desired class, to mix the module in the class. It is possible to mix more than one module in a class, thus it looks very similar to multiple inheritance. In the example from figure \ref{fig:diamond}, the superclasses could be two modules that implement the different move behaviours. But even if Ruby uses mixins instead of multiple inheritance, the nameclash problem persists. Ruby resolves it automatically by overriding the previous definition, thus it is important for the programmer to be aware of this and give another name to one of the definition if the two method definitions are needed.

@@

Polymorphism is a key feature in object-oriented programming languages. Both Ruby and Eiffel support this feature but in a very different conceptual way. Eiffel supports subtype polymorphism, it means that Eiffel allows polymorphism only for the types that have a superclass in common. Thus inheritance is primordial for  subtype polymorphism. 
% http://en.wikipedia.org/wiki/Duck_typing
In Ruby it is also possible to achieve it using inheritance, but this is more a consequence of the mechanism that permits Ruby to support polymorphism. Ruby is dynamically typed and supports a special style of typing namely, \emph{duck typing}. Duck typing focusses on the methods of an object instead of its type. If the method is supported by the object it will be called whatever the type or output is. If the method call is not supported, then a run-time error is returned. Duck typing is the concept used for polymorphism in Ruby allows thus polymorphism without inheriting from a superclass. It is trivial why polymorphism also works with inheritance in Ruby, classes that inherit from a superclass inherit its methods and duck typing focusses on the presence of methods.
% http://blogs.msdn.com/b/ericwhite/archive/2008/10/29/anders-hejlsberg-introduces-c-4-0-at-pdc-2008.aspx
So even if both languages support polymorphism, their approach is completely different. Polymorphic calls are dependent of the type of objects in Eiffel and in Ruby they are dependent of the presence of the method. From a software engineering point of view it is logical that Eiffel focusses on the subtype polymorphism. First it is statically typed, thus there should be a specific type declared with the variable or parameter. However, this could be resolved with a keyword that instructs the compiler that a variable should be dynamically typed. Second and most importantly, this sort of solution is not in the philosophy of the Eiffel language. One of the goals of Eiffel is to produce software that is reliable and maintainable. If duck typing should be adapted in Eiffel then it would be only reliable if the programmers know exactly which type of objects will passed to the methods. The maintainability of the software would also be tricky, imagine that a method call changes of name in one of the classes, with duck typing it would not be clear that this method was used in a polymorphic call and by changing its name the programmer introduces errors that can be tedious to resolve. Thus it is important to be aware of the methods that are used in a polymorphic context when a language supports duck typing. Subtype polymorphism in a statically typed language is a much more reliable mechanism because the polymorphic calls are only possible for a restricted set of types and eliminate a lot of possible failures. It is also more easily maintained, because it is checked at compile time that every class implements the method. In Ruby's case, it is also logical that duck typing is used and not subtype polymorphism. First, subtype polymorphism is implicit with duck typing. Second, Ruby is dynamically typed, which means the types are known at run-time. This implies that there would be no secure way to ensure that the method call is applied on a particular class hierarchy. Perhaps it could be achieved with a kind of type cast for the desired superclass. But then it would still return a run-time error when the types are not correct, like it does when the method is not defined for an object. This is much more restrictive and goes also against the philosophy of Ruby which is simplicity.

@@

If there is one language specific feature that is essential in Eiffel, then this concept is \emph{Design by Contract}. The idea is that every system has interacting components and that their cooperation should follow some strict specifications (the contract) that settle the obligations and benefits for both client and supplier. The obligations have to be satisfied before feature calls. They are called \emph{preconditions} and are introduced by the \emph{require} keyword. The benefits describe what the result should be if the precondition was met. Benefits are thus \emph{postconditions} and can be specified using the \emph{ensure} keyword. Every contract also include \emph{class invariants}, which are conditions that have to be ensured during the lifetime of an object, including at its creation. Class invariants are specified after the \emph{invariant} keyword. These are the three main categories of contracts and are implemented using assertions. Each assertion may be tagged, it is not mandatory and does not influence the contract but it is helpful for debugging and provides extra documentation. Since assertions are boolean expressions, it is possible to formulate them in function calls. This enables to express more complex conditions.

@@

There exist libraries in many programming languages that offers Design by Contract, even for \href{https://github.com/egonSchiele/contracts.ruby}{Ruby}. A basic implementation of Design by Contract in Ruby \ref{rb:contract} has been implemented for this paper. It uses reflection in order to get information about the variables and metaprogramming for evaluating the expressions passed to the require, ensure and invariant clauses.

@@

There are patterns that check if a variable is void-safe. The Certified Attachment Pattern (CAP) checks if a local variables or formal parameters is void. The attached syntax takes a step further. It is another sort of CAP that checks if the object is attached and provides a safe access to the objects that are attached as class attributes. Eiffel introduced two kind of types in order to assure the void-safeness of the software:

@@

This is a feature that improves the reliability of the produced software and shows again that Eiffel's primary concern is to enhance the software quality. In Ruby there is nothing like void-safety. Void safety is achieved at compile time and Ruby is an interpreted language, thus there is no way to make it void safe except by checking if the object is the nil object.
\subsection{Ruby}
Ruby is known for being very object-oriented and for its simplicity. It is a programming language that is very flexible like duck-typing already attested in previous section.
\subsubsection{Open Classes}
In Ruby it is always possible to add new methods to an existing class, the class definitions can at any moment be opened for modifications. Even built-in classes are can be adapted. Adding or modifying content at run-time to an already existing class definition without altering the source code is known as monkey patching. % wikipedia monkey patching
\\
\\
Monkey patching can be useful for different applications like extending or modifying the behaviour of an object at run-time from a third-party software without changing the source code. This enables for example to reuse existing code and adapt it in different files with different behaviours.
\\
\\
In first instance this feature seems really helpful because it gives the possibility to the developer to adapt code when needed. For example, as workaround to a bug by implementing a specific solution in this part of the code. But with great power comes great responsibility. Monkey patching can lead to bugs that break the code. If the source code changes, it is possible that the patch behaves differently because it makes assumptions on the code that do not hold anymore. If there are different patches for the same method it is possible that they enter in conflict and one will override the other. It is possible to save the implementation of a method by renaming it with an alias. It is also important to document the patches and method that are patched, this can prevent confusion if one forgets or does not know about the patch and does not understand why the code does not act like expected.
\\
\\
It is very important to note that open classes are in conflict with the principle of encapsulation. Every method can be accessed with open classes, even private methods and they can  be changed into public ones. It is thus really important to not abuse of this feature. Monkey-patching should only be used on third-party software if other concepts like inheritance fail.
\subsubsection{Meta-class model}

@@

%DESIGN BY CONTRACT CODE
\lstinputlisting[label=rb:contract,language=Ruby,caption={Design by Contract in Ruby}]{code_Ruby/contract.rb}
\lstinputlisting[label=e:contract,language=Eiffel,caption={Design by Contract in Eiffel}]{code_Eiffel/contract.e}

%OPEN CLASSES CODE
\lstinputlisting[label=rb:openclasses,language=Ruby,caption={Open classes in Ruby}]{code_Ruby/openclass.rb}
